Continuous gravitational waves in the lab: recovering audio signals with a table-top optical microphone

### 2020-06-16 paper/ifo-tableTopGWs.tex
going through a few more comments from Andrew

The basic physics of gravitational-wave detection can be demonstrated with a table-top Michelson interferometer, an optical configuration commonly used in undergraduate labs. 
A simple Michelson interferometer comprises of a laser, a beamsplitter, and two mirrors. 
Laser light is split by a beamsplitter into two arms at right angles to each other. 
At the end of each arm is a mirror which reflects the beams back to the beamsplitter where they recombine. 
The combination of the two beams produces an interference pattern which is dependent on the relative distance traveled by each of the beams. 
A change in arm length results in a change in the interference pattern.

Gravitational-wave detectors are large, complex interferometers.
The LIGO and Virgo detectors have $4\,{\rm km}$ and $3\,{\rm km}$ arms respectively.
They are sensitive to miniscule changes in distance, for example the first detection of a binary black hole merger had a peak gravitational-wave strain of $10^{-21}$ ~\cite{GW150914}.
However, their design is based on the Michelson interferometer. 
Detection and analysis concepts can be demonstrated with table-top optical equipment.

A commonly-used analogy in explaining gravitational-wave science is sound. 
Signals of binary black hole and neutron star mergers can be converted into audio as an aid to explanations~\cite{SoundsOfSpaceTime:online,BlackHoleHunter:online} (however it is important to note that gravitational waves are not sound).  
Sound vibrations provide a simple means to move the components of a demonstration interferometer. 
The vibrations move the mirrors, changing the length of the interferometer arms, and the resulting interference pattern. 
We use sound throughout this work as a means of simulating gravitational-wave-like signals and the data analysis and signal processing techniques used to detect them.

\begin{figure}
% 	\includegraphics[width=0.49\textwidth]{figures/ifo_schematic_webcam.pdf}
	\includegraphics[width=0.49\textwidth]{figures/ifo_schematic_webcam_edit.pdf}
	\caption{\label{fig:ifo_schematic_webcam}
Schematic of the table-top Michelson interferometer. 
The laser (top-left) is a Class 2 $532\,{\rm nm}$ laser~\cite{ThorLabsIFO}. 
The beam is first reflected from a mirror at a right angle and the focused at a converging lens before the beamsplitter which is $50$/$50$ in reflection and transmission. 
The split beams are reflected from mirrors (M1 and M2) at the ends of the interferometer arms and are recombined at the beamsplitter. 
The beam is enlarged by a diverging lens after the beamsplitter. 
The interference pattern is projected onto a screen and recorded using a commercial webcam connected to a personal computer (PC). 
A speaker fixed to the back of M2 provides the signal input from the PC.  
%Michelson interferometer schematic with interference pattern recorded by a commercial webcam. The laser (top-left) is a Class 2 $532\,{\rm nm}$ laser~\cite{ThorLabsIFO}. The beam is first reflected from a mirror a right angle (to save space on the optical breadboard) and then focused at a converging lens before the $50$/$50$ beamsplitter. The split beam is reflected from mirrors M1 and M2 at the ends of the interferometer arms and recombined at the beamsplitter. The interference pattern is enlarged by a diverging lens after the beamsplitter and projected onto a screen. A commertial speaker is fixed to the back of M2, as the signal input. The interferometer arms lengths are roughly $3\,{\rm cm}$ different and the entire interferometer (without the screen) fits within the dimensions of an A4 piece of paper. We did not include a compensating plate with the beamsplitter, as is often done.} 
    }
\end{figure}

The Michelson interferometer configuration used in this work is shown schematically in Fig.~\ref{fig:ifo_schematic_webcam}.
It is constructed on a $450 \times 450 \,{\rm mm} $ optical breadboard. 
The laser (top-left) has a wavelength of $532\,{\rm nm}$.
In the configuration shown in Fig.~\ref{fig:ifo_schematic_webcam} the laser beam is first reflected from a mirror which turns the beam $90^{\circ}$ (to save space on the optical breadboard). 
The beam passes through a a converging lens and then to the beamsplitter, which is $50$/$50$ reflection and transmission. 
The beams are reflected by mirrors at the end of each of the two arms ($\approx 7.5\,{\rm cm}$ and $\approx 10.0\,{\rm cm}$ long) and are recombined at the beamsplitter. 
The resulting interference pattern is projected onto a screen and recorded by a webcam.

@@

\begin{figure}
 \begin{center}
  \includegraphics[width=.45\textwidth]{figures/webcam_still0_crop.pdf}
 \end{center}
 \caption{\label{fig:interference_pattern}
 Interference pattern from Michelson interferometer as viewed by webcam from off-centre (hence why the pattern appears non-circular), red dot indicates point where intensity time series was measured from. For scale, the central fringe is approximately $5\,{\rm mm}$ across.}
\end{figure}

A $0.5\,{\rm W}$ speaker is fixed to the back of one of the mirrors (M2 in Fig.~\ref{fig:ifo_schematic_webcam}), providing the audio source for the set-up. 
The speaker is part of a commercial $3.5\,{\rm mm}$-jack, USB powered pair of speakers and was driven by a PC. 
It is attached to the back of the mirror using commercial adhesive putty. 
The other speaker in the pair is kept face-down and as far away from the interferometer as the cables allow to reduce it acting as a second noise source.

As the speaker deforms to play sound, the mirror moves producing a change in arm length related to some coupling function of the amplitude and frequency of the audio. 
The motion of the interference fringes follows the motion of the mirror, and therefore the speaker, at the same frequency. 
If the fringe motion is small, then changes in the intensity at some point on the screen follows the frequency of the injected audio but with amplitude given by a transfer function that accounts for the coupling and resonance of the speaker-mirror system.
If the fringe motion is large enough for multiple fringes to pass over the measured point in a single speaker deformation, then over-counting artificially raises the measured frequency. As such, any motion of the fringes must be kept small by playing sound softly through the speaker.
For later sections, we derive the non-linear relation between the separation of the mirrors and the intensity of the pattern at some point on the screen, this can be found in Appendix~\ref{app:intensity_derivation}. %\han{move over!}

Data can be recorded via a webcam (as shown in Fig.~\ref{fig:ifo_schematic_webcam}) or photo-diode. 
In Sections~\ref{sec:single_tone} and~\ref{sec:viterbi_wandering} a webcam is used. 
In Section~\ref{sec:optical_microphone}, which considers more complex audio, a photodiode is used as it provides a higher sample rate than the webcam.

>>>>

The basic physics of gravitational-wave detection can be demonstrated with a table-top Michelson interferometer, an optical configuration commonly used in undergraduate labs. 
A simple Michelson interferometer comprises a laser, a beamsplitter, and two mirrors. 
Laser light is split by a beamsplitter into two arms at right angles. 
At the end of each arm is a mirror which reflects the beams back to the beamsplitter where they recombine. 
The combination of the two beams produces an interference pattern which is dependent on the relative distance travelled by each of the beams. 
A change in arm length results in a change in the interference pattern.

Gravitational-wave detectors are large, complex interferometers.
The LIGO and Virgo detectors have $4\,{\rm km}$ and $3\,{\rm km}$ arms respectively.
They are sensitive to miniscule changes in distance, for example the first detection of a binary black hole merger had a peak gravitational-wave strain of $10^{-21}$ ~\cite{GW150914}.
However, their design is based fundamentally on the Michelson interferometer. 
Detection and analysis concepts can be demonstrated with table-top optical equipment.

A commonly-used analogy in explaining gravitational-wave science is sound. 
Signals of binary black hole and neutron star mergers can be converted into audio as an aid to explanations~\cite{SoundsOfSpaceTime:online,BlackHoleHunter:online}~\footnote{When introducing the acoustic analogy to lay audiences, it ins important to emphasize that gravitational waves are not sound. For example, gravitational waves can propagate in a vacuum, whereas sound cannot.}.
Sound vibrations provide a simple means to move the components of a demonstration interferometer. 
The vibrations move the mirrors, changing the length of the interferometer arms, and the resulting interference pattern. 
We use sound throughout this work as a means of simulating gravitational-wave-like signals and the data analysis and signal processing techniques used to detect them.

\begin{figure}
% 	\includegraphics[width=0.49\textwidth]{figures/ifo_schematic_webcam.pdf}
	\includegraphics[width=0.49\textwidth]{figures/ifo_schematic_webcam_edit.pdf}
	\caption{\label{fig:ifo_schematic_webcam}
Schematic of the table-top Michelson interferometer. 
The laser (top-left) is a Class 2 $532\,{\rm nm}$ laser~\cite{ThorLabsIFO}. 
The beam is first reflected from a mirror at a right angle and then focused at a converging lens before the beamsplitter which is $50$/$50$ in reflection and transmission. 
The split beams are reflected from mirrors (M1 and M2) at the ends of the interferometer arms and are recombined at the beamsplitter. 
The beam is enlarged by a diverging lens after the beamsplitter. 
The interference pattern is projected onto a screen and recorded using a commercial webcam connected to a personal computer (PC). 
A speaker fixed to the back of M2 provides the signal input from the PC.  
%Michelson interferometer schematic with interference pattern recorded by a commercial webcam. The laser (top-left) is a Class 2 $532\,{\rm nm}$ laser~\cite{ThorLabsIFO}. The beam is first reflected from a mirror a right angle (to save space on the optical breadboard) and then focused at a converging lens before the $50$/$50$ beamsplitter. The split beam is reflected from mirrors M1 and M2 at the ends of the interferometer arms and recombined at the beamsplitter. The interference pattern is enlarged by a diverging lens after the beamsplitter and projected onto a screen. A commertial speaker is fixed to the back of M2, as the signal input. The interferometer arms lengths are roughly $3\,{\rm cm}$ different and the entire interferometer (without the screen) fits within the dimensions of an A4 piece of paper. We did not include a compensating plate with the beamsplitter, as is often done.} 
    }
\end{figure}

The Michelson interferometer configuration used in this work is shown schematically in Fig.~\ref{fig:ifo_schematic_webcam}.
It is constructed on a $450 \times 450 \,{\rm mm} $ optical breadboard. 
The laser (top-left) has a wavelength of $532\,{\rm nm}$.
In the configuration shown in Fig.~\ref{fig:ifo_schematic_webcam} the laser beam is first reflected from a mirror which turns the beam $90^{\circ}$ (to save space on the optical breadboard). 
The beam passes through a converging lens and then to the beamsplitter, which is $50$/$50$ reflection and transmission. 
The beams are reflected by mirrors at the end of each of the two arms ($\approx 7.5\,{\rm cm}$ and $\approx 10.0\,{\rm cm}$ long) and are recombined at the beamsplitter. 
The resulting interference pattern is projected onto a screen and recorded by a webcam.

@@

\begin{figure}
 \begin{center}
  \includegraphics[width=.45\textwidth]{figures/webcam_still0_crop.pdf}
 \end{center}
 \caption{\label{fig:interference_pattern}
Michelson interference pattern as viewed by an off-axis webcam (which is why the pattern is elliptical). 
The red dot marks where the intensity time series is measured. 
The central bright fringe of the interference pattern is approximately $5\,{\rm mm}$ in diameter. 
% Interference pattern from Michelson interferometer as viewed by webcam from off-centre (hence why the pattern appears non-circular), red dot indicates point where intensity time series was measured from. For scale, the central fringe is approximately $5\,{\rm mm}$ across.
}
\end{figure}

A $0.5\,{\rm W}$ speaker is fixed to the back of one of the mirrors (M2 in Fig.~\ref{fig:ifo_schematic_webcam}), providing a controllable source of vibrations. 
The speaker is one half of a pair of commercial, USB-powered speakers fed by a $3.5\,{\rm mm}$ jack and driven by a PC. 
It is attached to the back of the mirror using commercial adhesive putty. 
The other speaker in the pair is kept face-down and as far away from the interferometer as the cables to reduce its impact as a second source of vibrations.

%As the speaker deforms to play sound, the mirror moves producing a change in arm length related to some coupling function of the amplitude and frequency of the audio. 
The motion of the interference fringes follows the motion of the mirror, and therefore the speaker. 
If the fringe motion is small, then changes in the intensity at some point on the screen occur at the same frequency as the injected audio but with amplitude given by a transfer function that accounts for the coupling and resonance of the speaker-mirror system.
If the fringe motion is large enough for multiple fringes to pass over the measured point in a single speaker deformation, then over-counting artificially raises the measured frequency. As such, any motion of the fringes must be kept small by playing sound softly through the speaker.
Even without over-counting, large fringe motions introduce a nonlinear relation between the intensity fluctuations and injected audio, leading to troublesome - but physically interesting! - phenomena like frequency doubling.
We derive the non-linear relation between the separation of the mirrors and the intensity of the pattern at some point on the screen in Appendix~\ref{app:intensity_derivation}.

Data can be recorded via a webcam (as shown in Fig.~\ref{fig:ifo_schematic_webcam}) or photo-diode. 
In Sections~\ref{sec:single_tone} and~\ref{sec:viterbi_wandering} a webcam is used. 
In Section~\ref{sec:optical_microphone}, which considers more complex audio, we turn to a photodiode because it offers a higher sample rate than the webcam.

### 2020-06-16 paper/ifo-wanderingTone.tex
going through a few more comments from Andrew

% Tracking tones with the Viterbi algorithm
% explaination and successful application
% lean on the continuous gravitational waves angle
Continuous wave signals may wander slowly in frequency over time. 
%As also mentioned in Section~\ref{sec:introduction}, continuous wave searches often look for signals that wander slowly in frequency~\cite{ScoX1O2Viterbi:2019}. 
The audio analogue of this is a tone that wanders in frequency; a note that changes pitch. 
%Unlike the single tone described in Section~\ref{sec:single_tone}, finding the signal isn’t as simple as searching for a peak in the spectrum for the full timeseries data.%observing run.

The analysis techniques used here are inspired by those used to search for continuous waves from spinning neutron in LIGO and Virgo data~\cite{SuvorovaEtAl:2016,SuvorovaEtAl:2017}.
Continuous wave searches are performed on long datasets (months to years in duration). 
The frequency of gravitational-wave emission wanders significantly over the observation period.

One method to search for a wandering signal is to split the data into several shorter segments which are analysed separately initially. 
The resulting segments are combined to form a grid in time and frequency (spectogram). 
The value at each spectrogram point determines the likelihood that a signal is present at a particular time and frequency. 
In gravitational-wave analysis, a detection statistic is used to calculate the likelihood of a signal (the statistic used depends on the type of target~\cite{JKS:1998,SuvorovaEtAl:2017}). 
In this work we take a Fourier transform of each time segment.
The resulting spectrogram gives the Fourier amplitude for each frequency and time bin. 
The task is then to find a wandering frequency signal through the spectogram.

The Viterbi algorithm~\cite{Viterbi:1967} is an efficient method to find the most likely path of a signal given an observation sequence. 
It is used in continuous wave searches for a range of astrophysical targets~\cite{ScoX1O2Viterbi:2019,ScoX1ViterbiO1:2017,MillhouseStrangMelatos:2020,PostMergerRemnantSearch:2019,SunEtAlSNR:2018,viterbi_application}.

In Section~\ref{sec:viterbi} we review the Viterbi algorithm and its application to this work and in Section~\ref{sec:wanderingResults} we present our results for recovery of a wandering audio frequency.

@@

\subsection{The Viterbi algorithm}
\label{sec:viterbi}
%Here we overview the Viterbi algorithm and describe its usage in this work. 
%Thinking about this - should we add some maths?
%Our starting point is the spectrogram grid in time $t$ and frequency $f$. 
%Each element in the grid contains the Fourier amplitude for a particular $f$ and $t$. 
%Our objective is to find the most likely through the grid which maximises the summed Fourier amplitudes along the path. 
%In our implementation we sum the log of the normalised 
%We label the time-frequency grid points as $t_i$ and $f_j$ respectively, where $i=1,2,...N_t$ and $j=1,2,...,N_f$ (and $N_t$ and $N_f$ are the total number of time and frequency bins respectively). 
%Starting at $t_{i=2}$, the algorithm iterates through $t_i$, at each stage finding the best path to every $f_j$ from $t_{i=0}$.

@@

% moved from above
The method requires a limit for the allowed frequency change over time, usually informed by a physical model. 
In this work the frequency $f$ path can move in one of three ways at each time bin: a) up one frequency bin; b) stay in the same frequency bin; or c) down one frequency bin. 
We set the frequency bin width to be $\Delta f = 0.3\,{\rm Hz}$ which corresponds to a maximum frequency wander of $0.11\,{\rm Hz}{\rm s}^{-1}$.
%In this work, the frequency wander limit is set to $0.11\,{\rm Hz} {\rm s}^{-1}$ which corresponds to one frequency bin per every time bin. 
In some cases, we see this proves poorly adjusted to the signal.

\subsection{Results}
\label{sec:wanderingResults}

A wandering frequency signal is played by the speaker and the output recorded via webcam as in Section~\ref{sec:single_tone}.
Note that when creating a wandering frequency signal, the phase change must be integrated up over time, as detailed in Appendix~\ref{app:phase_gotcha}.

\begin{figure*}
	\includegraphics[width=\textwidth]{figures/expt_overlay_2_viterbi_test_webcam.pdf}
	\caption{\label{fig:viterbi_overlay}
Recovery of a wandering tone. 
The spectrogram shows the observed frequency amplitdue at each time-frequency step. 
The overlaid pink-dot and white-cross markers show the injected signal and recovered Viterbi path respetively. 
On the left, before $\sim 15\,{\rm s}$, the signal changes frequency too quickly for the Viterbi algorithm to recover. 
At $150\,{\rm s}$ the data appears anomalous, which may be due to some background noise. }
\end{figure*}

Our results are shown in Fig.~\ref{fig:viterbi_overlay}, where the heatmap shows the spectrogram of the observed signal. 
The overlaid pick-dot and white-cross markers show the injected signal and recovered Viterbi path respectively. 
The Viterbi path provides a good match to the injected frequency. 
%To the eye, the Viterbi algorithm easily recovers the injected frequency over time due to the high signal-to-noise ratio (SNR) of this experiment \han{What was the SNR? / can we quantify this?}.
Quantitatively, the Viterbi path is within one frequency bin ($\approx 0.3\,{\rm Hz}$) of the injected signal for $94\%$ of the time. 
%The Viterbi path stays within one frequency bin (approximately $0.3\,{\rm Hz}$) of the injected tone for $94\%$ of the run. 
Initially the injected signal moves faster then the frequency wander limit, leading to the discrepancy for $t\lesssim 10\,{\rm s}$. 
%Note that the signal initially moves faster than the frequency wander limit, causing a discrepancy between the injected and recovered signals. 
%To fix this, the limit should be doubled, although this would then cause the recovered path to stray more often. 
There is also an anomaly at $150\,{\rm s}$, which is likely due to a disturbance around the interferometer (such as someone walking nearby).
%There also appears to be an anomaly at $150\,{\rm s}$, likely from a disturbance (such as someone walking) around the interferometer.

>>>>

% Tracking tones with the Viterbi algorithm
% explaination and successful application
% lean on the continuous gravitational waves angle
Continuous gravitational wave signals may wander slowly in frequency over time, due to stochastic internal processes in the superfluid interior of isolated neutron stars~\cite{MelatosDouglassSimula:2015,Jones:2010} or stochastic accretion flows in binary neutron stars~\cite{BildstenTB:1998}. 
%As also mentioned in Section~\ref{sec:introduction}, continuous wave searches often look for signals that wander slowly in frequency~\cite{ScoX1O2Viterbi:2019}. 
The audio analogue of this is a tone that wanders in frequency; a note that changes pitch. 
%Unlike the single tone described in Section~\ref{sec:single_tone}, finding the signal isn’t as simple as searching for a peak in the spectrum for the full timeseries data.%observing run.

The analysis techniques used here are inspired by those used to search for continuous waves from spinning neutron in LIGO and Virgo data~\cite{SuvorovaEtAl:2016,SuvorovaEtAl:2017}.
Continuous wave searches are performed on long datasets (months to years in duration). 
The frequency of gravitational-wave emission wanders significantly over the observation period. 
In this context, ``significantly'' means ``across multiple frequency bins'', where the typical width of a frequency bin is the reciprocal of the total observing time~\cite{JKS:1998,ScoX1O2Viterbi:2019}

One method to search for a wandering signal is to split the data into several shorter segments which are analysed individually. 
The detection statistics in every segment are combined to form a grid in time and frequency (spectogram). 
The value at each spectrogram point determines the likelihood that a signal is present at a particular time and frequency. 
In gravitational-wave data analysis, a detection statistic is used to calculate the likelihood of a signal given the antenna beam pattern of the detector, which varies with as the Earth rotates and orbits the Sun~\cite{JKS:1998}.
The detection statistic used depends on the type of target~\cite{JKS:1998,SuvorovaEtAl:2017}). 
In this work the detection statistic is the amplitude (modulus) of the Fourier transform of the data at each time segement, which maximizes the likelihood of detecting a sinusiod in Gaussian noise, as described in Appendix\han{to do}.
The resulting spectrogram gives the Fourier amplitude for each frequency and time bin. 
The task is then to find the most likely wandering frequency signal through the spectogram. 
The Viterbi algorithm~\cite{Viterbi:1967} is an efficient method to find the most likely path of a signal given an observation sequence. 
It is used in continuous wave searches for a range of astrophysical targets~\cite{ScoX1O2Viterbi:2019,ScoX1ViterbiO1:2017,MillhouseStrangMelatos:2020,PostMergerRemnantSearch:2019,SunEtAlSNR:2018,viterbi_application}.

In Section~\ref{sec:viterbi} we review the Viterbi algorithm and its application to this work. 
In Section~\ref{sec:wanderingResults} we present our results for recovery of a wandering audio frequency.

@@

\subsection{The Viterbi algorithm}
\label{sec:viterbi}

\begin{figure}
\includegraphics[width=0.49\textwidth]{/home/hannahm/Documents/viterbi/quickTest/circlesTest}
\caption{\label{fig:viterbi}
A schematic diagram of the Viterbi algorithm. 
Each box represents an element in the time-frequency grid starting from $t_0$ on the left and ending at $t_4$ on the right. 
The color of the box indicates the likelihood that a signal is present in the box, where a lighter color corresponds to higher likelihood. 
Each box has a single path leading to it indicated by the dotted lines. 
The Viterbi path is highlighted in orange. 
At each time step, the paths can move up one $f$ box, move down one $f$ box, or stay in the same $f$ box. 
See the main text for a full description. 
}
\end{figure}

%Here we overview the Viterbi algorithm and describe its usage in this work. 
%Thinking about this 
Our starting point is the spectrogram grid indexed by time $t$ and frequency $f$ where there are $N_t$ and $N_f$ time and frequency bins respectively as shown in Fig.~\ref{fig:viterbi}.
We label the elements of the grid as $t_i$ and $f_j$ with $i=0,1,2,...N_t$ and $j=0,1,2,...,N_f$. 
The value in each grid element is a detection statistic based on the time series data observed. 
It indicates the likelihood that a signal is present in the element. 
We label the normalised Fourier amplitude as $F(t_i,f_j)$. 
%In continuous wave searches the detection statistic depends on the type of gravitational wave source. 
%In this work we use the normalised Fourier amplitude, which we label $F(t_i,f_j)$.

Our objective is to find the most likely path through the grid given the data and any probabilistic constraints on how the frequency wanders from $t_i$ to $t_{i+1}$. 
In continuous wave searches, a physical model is used to decide how much the frequency of the signal is allowed to wander at each time step, which depends on the type of target. 
In this work, at each time step we allow the frequency to either a) stay in the same bin; b) move up a single frequency bin; or c) move down a single frequency bin. 
The probability of transitioning from $f_j$ at $t_i$ to any of $\{f_{j+1},f_j,f_{j-1}\}$ at $t_{i+1}$ is called the transition probability $A(f_k,f_m)$. 
In this work the three options have equal probability $A(f_k,f_m)=1/3$ for $k=m+1,m,m-1$ and $A(f_k,f_m)=0$ otherwise

A path through the grid has a likelihood dependent on the detection statistic (the observation) and the transition probability.
If the starting probabilities are all equal, i.e. $Pr[f(t_0)] = 1/N_f$, the the probability of the frequency equalling $f(t_{i+1})$ at $t = t_{i+1}$ is defined recursively by
%If there is equal probability of starting $f_j$ at $t=t_0$, the probability of path $Q$ given the observation $O$ is 
\begin{eqnarray}
%P(Q|O) =& A(f_j) F(t=N_t,f_j), \nonumber \\ &~~ \times \dots, A(f_j) F (t=t_0,f_j).
Pr[f(t_{i+1})=f_j] =~& F[t_{i+1},f(t_{i+1})] \nonumber \\
                     &*A[f(t_{i+1}),f(t_i)]  \nonumber \\
                     & *Pr[f(t_i)].
\label{eqn:recursiveViterbi}
\end{eqnarray}
The sequence $f(t_0),f(t_1),\dots,f(t_{N_t})$ that maximizes $Pr[f(t_{N_t})] = f_j]$ is the optimal Viterbi path terminating in the frequency bin $f_j$. 
We can the maximize the latter quantity over $0 \leq j \leq N_f$ to find the optimal Viterbi path overall, i.e. terminating in any frequency bin.

The Viterbi algorithm is an efficient method to find the Viterbi path. 
Here we describe the algorithm as psudeo-code referring to the schematic in Fig.~\ref{fig:viterbi}, where the circles indicate the elements of the spectrogram grid (lighter colours correspond to higher likelihood). 
The implementation used in this work is available online (see Appendix~\ref{app:code}).
\begin{enumerate}
\item Starting at $t_1$, each $f_j$ element has three possible paths it could originate from at $t_0$, indicated by the lines in Fig.~\ref{fig:viterbi}. 
At each $f_j$ element, the highest $F[t_0,f_j] A[f(t_0),f(t_1)]$ from the previous time step is selected, shown by the solid lines in Fig.~\ref{fig:viterbi}. 
For example for the element labelled (a), the highest value is directly behind it, therefore the transition from $\{t_0,f_0\}$ to $\{t_1,f_0\}$ is selected. 
Each element has now selected the best path from the previous time step. 
In Fig.~\ref{fig:viterbi}, black lines show the best paths and grey lines show all other paths. 
\item Moving to $t_2$, the path with the highest value of the recursive relation in equation~\ref{eqn:recursiveViterbi} is selected for each $f_j$ as shown by the solid lines from $t_1$ to $t_2$ in Fig.~\ref{fig:viterbi}. 
\item Steps 2 is repeated until the end of the grid is reached. 
\item At $t=t_{N_t}$, each $f_j$ has a value $Pr[f(t_{N_t})=f_j]$ (equation~\ref{eqn:recursiveViterbi}). 
The highest likelihood path for each $f_j$ can be seen by tracing the solid lines from right to left in Fig.~\ref{fig:viterbi}. 
The $f_j$ with the maximum $Pr[f(t_{N_t})=f_j]$ is labelled (b) in Fig.~\ref{fig:viterbi}. 
\item Tracing the p
At the end of the grid, each $f_j$ has a unique path associated with it (seen by tracing the dotted lines from right to left in Fig.~\ref{fig:viterbi}). The element with the highest accumulated probability is selected, in the case of Fig.~\ref{fig:viterbi}, this is labelled (b). 
\item Tracing the path from right to left in Fig.~\ref{fig:viterbi}, we see that the path ending at (b) started at (c). This is the Viterbi path, which is highlighted in orange in Fig.~\ref{fig:viterbi}. 
\end{enumerate}

\begin{comment}
The route through the grid which maximises the overall probability

Constraints are set on the allowable paths. 
In continuous wave searches, a physical model is used to decide how much the frequency can wander at each timestep, depends on the type of target. 
In this work we allow the frequency to either a) stay in the same bin; b) move up a single frequency bin; or c) move down a single frequency bin. 
We label the probability of transitioning from $f_j$ at $t_{i}$ to $f_{j+1,j,j-1}$ at $t_{i+1}$ as $A_{i,j}$. 
In this work the three transition option have equal probability.

Our objective is now to find the most likely path through the grid given the data and transition constraints. 
This is the route which maximises the overall probability, which we call the `Viterbi path'.
In practice we sum the logarithm of the Fourier amplitude $\log O_{i,j}$ along the path.

The Viterbi algorithm begins at $t_{i=1}$. 
The algorithm selects the highest probability route for each $f_j$. 
\begin{equation}
A_{i,j} O_{i,j+1}
A_{i,j} O_{i,j}
A_{i,j} O_{i,j-1}
\end{equation}
At $t_{i=2}$ it then selects the best route to this point based on the accumulated probability along the paths so far. 
For example, at

(a.) The grid point labelled (a) has $f_{j=3}$ and $t_{i=1}$. 
Looking back at the previous time, there are three options for where the path could have started from at $t_{i=0}$. These are $f_{j=4}$, $f_{j=3}$, or $f_{j=2}$. 
The algorithm selects $f_{j=3}$ as it has the highest value. 
(b.) The grid point labelled (b.) is not the best starting position for any of the cells at $t_{i=1}$, no paths starting from this position will be considered from now on.

The probability of a path is given by the accumulated probability along the path. 
\begin{equation}

\end{equation}

The value of each element in the grid is the logarithm of the normalised Fourier amplitude, which we label $O_{i,j}$.

Our objective is to find the most likely path through the grid which maximises the , the `best path'. 
In practice, we sum the logarithm of the Fourier amplitudes along the path.

We label the time-frequency grid points as $t_i$ and $f_j$ respectively, where $i=1,2,...N_t$ and $j=1,2,...,N_f$ (and $N_t$ and $N_f$ are the total number of time and frequency bins respectively). 
Starting at $t_{i=2}$, the algorithm iterates calculates the best path to each $f_j$ from $t_{i=1}$ to $t_=2$.  through $t_i$, at each stage finding the best path to every $f_j$ from $t_{i=0}$.

@@

% moved from above
The method requires a limit for the allowed frequency change over time, usually informed by a physical model. 
In this work the frequency $f$ path can move in one of three ways at each time bin: a) up one frequency bin; b) stay in the same frequency bin; or c) down one frequency bin. 
We set the frequency bin width to be $\Delta f = 0.3\,{\rm Hz}$ which corresponds to a maximum frequency wander of $0.11\,{\rm Hz}{\rm s}^{-1}$.
%In this work, the frequency wander limit is set to $0.11\,{\rm Hz} {\rm s}^{-1}$ which corresponds to one frequency bin per every time bin. 
In some cases, we see this proves poorly adjusted to the signal. 
\end{comment}

\subsection{Sample output}
\label{sec:wanderingResults}

To test the Viterbi algorithm, a signal with wandering frequency is played by the speaker and the output recorded via webcam as in Section~\ref{sec:single_tone}.
%Note that when creating a wandering frequency signal, the phase change must be integrated up over time, as detailed in Appendix~\ref{app:phase_gotcha}.

\begin{figure*}
	\includegraphics[width=\textwidth]{figures/expt_overlay_2_viterbi_test_webcam.pdf}
	\caption{\label{fig:viterbi_overlay}
Recovery of a wandering tone. 
The spectrogram shows the observed frequency amplitudue at each time-frequency step. 
The overlaid pink-dot and white-cross markers show the injected signal and recovered Viterbi path respectively. 
On the left, before $\sim 15\,{\rm s}$, the signal changes frequency too quickly for the Viterbi algorithm to recover. 
At $150\,{\rm s}$ the data appears anomalous, which may be due to some background noise. }
\end{figure*}

The results are shown in Fig.~\ref{fig:viterbi_overlay}, where the heatmap shows the spectrogram of the observed signal. 
The overlaid pink-dot and white-cross markers show the injected signal and recovered Viterbi path respectively maximized over all terminating frequency bins. 
The Viterbi path provides a good match to the injected frequency. 
%To the eye, the Viterbi algorithm easily recovers the injected frequency over time due to the high signal-to-noise ratio (SNR) of this experiment \han{What was the SNR? / can we quantify this?}.
Quantitatively, the Viterbi path is within one frequency bin ($\approx 0.3\,{\rm Hz}$) of the injected signal for $94\%$ of the time. 
%The Viterbi path stays within one frequency bin (approximately $0.3\,{\rm Hz}$) of the injected tone for $94\%$ of the run. 
Initially the injected signal moves faster then the frequency wander limit, leading to the discrepancy observed for $t\lesssim 10\,{\rm s}$. 
%Note that the signal initially moves faster than the frequency wander limit, causing a discrepancy between the injected and recovered signals. 
%To fix this, the limit should be doubled, although this would then cause the recovered path to stray more often. 
There is also an anomaly at $150\,{\rm s}$, which illustrates what happens when the interferometer is disturbed, e.g. when someone walks nearby. 
The Viterbi algorithm is robust against such disturbances, recovering the optimal path promptly, as Fig.~\ref{fig:viterbi_overlay} shows. 
%is likely due to a disturbance around the interferometer (such as someone walking nearby).
%There also appears to be an anomaly at $150\,{\rm s}$, likely from a disturbance (such as someone walking) around the interferometer.
